Sequential algorithms to split and merge terabyte-scale 3D images

### 2017-06-12 paper.tex
Added figure on buffer, added variants of cluster reads

@@

The following properties are straightforward applications of these
definitions:
\begin{itemize}
\item number of bytes per chunk: $\frac{Rb}{n}$
\item number of chunks that can fit in memory: $\frac{m.n}{Rb}$
\end{itemize}
For the sake of mathematical simplicity, the models below assume that
the reconstructed image and blocks are cubic. Under this assumption,
we have the following definitions and properties:
\begin{itemize}
\item Number of slices in a block: $\sqrt[3]{\frac{R}{n}}=d$. This is
  also the number of rows and columns in a slice and the number of
  voxel in a row or column.
\item Number of blocks required to fully cover a slice in the
  reconstructed image: $\sqrt[3]{n}^2=n^{\frac{2}{3}}$
  \item Number of full slices that can be reconstructed from $k$
    blocks: $\floor*{\frac{k}{n^{\frac{2}{3}}}}\sqrt[3]{\frac{R}{n}}$
\end{itemize}
\todo{Make a figure to illustrate the notations.}

@@

Algorithms~\ref{algo:naive-slices} and~\ref{algo:naive-blocks}
actually have very different performance even though blocks and slices
have identical sizes. Since slices are stored contiguously in the
reconstructed image, the number of seeks in
Algorithm~\ref{algo:naive-slices} is only $2n$:
\begin{equation}
  N_\mathrm{slices} = 2n 
\end{equation}
However,
Algorithm~\ref{algo:naive-blocks} has to do extra seeks for each row
in each slice of each block. That is, the total number of seeks
performed by Algorithm~\ref{algo:naive-blocks} is:
\begin{equation}
N_\mathrm{blocks} = 2n+nd^2 = n\left(1+\sqrt[3]{\frac{R}{n}}^2\right)
\end{equation}
In
practice, this difference could lead to a tremendous slowdown as we will show
later.

@@

As detailed in Algorithm~\ref{algo:cluster-reads}, cluster reads load
multiple blocks in memory, concatenate them in a buffer and write the
buffer sequentially in the reconstructed image. The buffer is a slightly complex
data structure, for instance an associative array or a Python
dictionary, capable of storing multiple disjoint sequences of
contiguous bytes without having to allocate memory for the bytes
between such sequences. Implementation details about the buffer are
given later in the document.

\begin{figure}
\centering
\def\svgwidth{0.3\columnwidth}
\input{figures/no-seek.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/d-seeks.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/d2-seeks.pdf_tex}
\caption{Left: complete slices and rows are reconstructed in memory:
  no seeking is required. Center: complete rows but incomplete slices
  are reconstructed: $d$ seeks are required. Right: incomplete rows
  are reconstructed: $d^2$ seeks are required.}
\label{fig:cluster-reads}
\end{figure}

\todo{a figure showing the buffer and how it
  fill up would help}.
%\note{More on defragmentation might be
%  needed here.}
\begin{algorithm}[h]
  \caption{Buffered merging of blocks with cluster reads}
  \label{algo:cluster-reads}
  \begin{algorithmic}
    \STATE list = sort blocks by increasing (x,y,z) values
    \STATE initialize buffer
    \STATE index = 0
    \WHILE{index\textless n}
      \WHILE{sizeof(buffer)\textless m}
        \STATE block = list[index]
        \STATE read block and insert it in buffer
        \STATE index += 1
        \ENDWHILE
       \STATE write buffer in reconstructed image
       \STATE clear buffer
    \ENDWHILE
  \end{algorithmic}
\end{algorithm}

The number of seeks performed by cluster reads depends on how blocks
loaded in memory arrange in the reconstructed image. In the best case,
complete contiguous slices of the reconstructed image can be assembled
in memory and written without seeking at all. In the worst case, some
rows in the reconstructed image are not covered by blocks loaded in
memory: $d^2$ seeks are then required, one for every partial row in
every partial slice. In the intermediary case, rows are complete but
some slices can only be partially reconstructed: $d$ seeks are then
required. In summary:
\begin{equation}
  N_\mathrm{cluster\_reads}= n + \ceil*{\frac{bR}{m}}f(m,n,R)
\end{equation}
where:
\begin{equation}
  f(m,n,R)=
  \begin{cases}
    d^2 & $if there is an incomplete row$\\
    d   & $if there is an incomplete slice but not incomplete row$\\
    1   & $otherwise$
\end{cases}
\end{equation}

@@

>>>>

\todo{Overlapping blocks. non-cubic blocks.}

@@

The following properties are straightforward applications of these
definitions:
\begin{itemize}
\item number of bytes per chunk: $\frac{Rb}{n}$
\item number of chunks that can fit in memory: $\frac{m.n}{Rb}$
\end{itemize}
For the sake of mathematical simplicity, the models below assume that
the reconstructed image and blocks are cubic. Under this assumption,
we have the following definitions and properties:
\begin{itemize}
\item Number of slices in a block: $\sqrt[3]{\frac{R}{n}}=d$. This is
  also the number of rows and columns in a slice and the number of
  voxel in a row or column.
\item Number of blocks required to fill a row of the reconstructed
  image: $\sqrt[3]{n}=\nu$.
\end{itemize}
\todo{Make a figure to illustrate the notations.}

@@

Algorithms~\ref{algo:naive-slices} and~\ref{algo:naive-blocks}
actually have very different performance even though blocks and slices
have identical sizes. Since slices are stored contiguously in the
reconstructed image, the number of seeks in
Algorithm~\ref{algo:naive-slices} is only $2n$:
\begin{equation}
  N_\mathrm{slices} = 2n 
\end{equation}
However,
Algorithm~\ref{algo:naive-blocks} has to do extra seeks for each row
in each slice of each block. That is, the total number of seeks
performed by Algorithm~\ref{algo:naive-blocks} is:
\begin{equation}
N_\mathrm{blocks} = n+nd^2 
\end{equation}
In
practice, this difference could lead to a tremendous slowdown as we will show
later.

@@

As detailed in Algorithm~\ref{algo:cluster-reads}, cluster reads load
multiple blocks in memory, concatenate them in a buffer and write the
buffer sequentially in the reconstructed image. The buffer is a
slightly complex data structure, for instance an associative array or
a Python dictionary, capable of storing multiple disjoint sequences of
contiguous bytes without having to allocate memory for the bytes
between such sequences. Figure~\ref{fig:cluster-reads-buffer}
illustrates how the buffer would fill up when two blocks of one slice
and three rows per slice are loaded. Implementation details about the
buffer are given later in the document.
\begin{figure}
\centering
\def\svgwidth{\columnwidth}
\input{figures/buffer-cluster-reads.pdf_tex}
\caption{Illustration of the buffer used in cluster reads. White
  portions are not allocated. [a,b] refers to block a, row b.}
\label{fig:cluster-reads-buffer}
\end{figure}
\begin{algorithm}[h]
  \caption{Buffered merging of blocks with cluster reads}
  \label{algo:cluster-reads}
  \begin{algorithmic}
    \STATE list = sort blocks by increasing (x,y,z) values
    \STATE initialize buffer
    \STATE index = 0
    \WHILE{index\textless n}
      \WHILE{sizeof(buffer)\textless m}
        \STATE block = list[index]
        \STATE read block and insert it in buffer
        \STATE index += 1
        \ENDWHILE
       \STATE write buffer in reconstructed image
       \STATE clear buffer
    \ENDWHILE
  \end{algorithmic}
\end{algorithm}

The number of seeks performed by cluster reads depends on how blocks
loaded in memory arrange in the reconstructed image. In the best case,
complete contiguous slices of the reconstructed image can be assembled
in memory and written without seeking at all. In the worst case, some
rows in the reconstructed image are not covered by blocks loaded in
memory: $d^2$ seeks are then required, one for every partial row in
every partial slice. In the intermediary case, rows are complete but
some slices can only be partially reconstructed: $d$ seeks are then
required. See also Figure~\ref{fig:cluster-reads}. In summary:
\begin{equation}
  N_\mathrm{cluster\_reads}= n + \ceil*{\frac{bR}{m}}f(m,n,R)
\end{equation}
where:
\begin{equation}
  f(m,n,R)=
  \begin{cases}
    d^2 & $if there is an incomplete row$\\
    d   & $if there is an incomplete slice$\\
        & $but not incomplete row$\\
    1   & $otherwise$
\end{cases}
\end{equation}
\begin{figure}
\centering
\def\svgwidth{0.3\columnwidth}
\input{figures/no-seek.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/d-seeks.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/d2-seeks.pdf_tex}
\caption{Left: complete slices and rows are reconstructed in memory:
  no seeking is required. Center: complete rows but incomplete slices
  are reconstructed: $d$ seeks are required. Right: incomplete rows
  are reconstructed: $d^2$ seeks are required.}
\label{fig:cluster-reads}
\end{figure}

If $\frac{Rb}{\sqrt[3]{n}}>m\geq\frac{Rb}{\sqrt[3]{n}^2}$, i.e., a
full row of blocks fits in memory but a full slice does not, then a
variant of cluster reads consists in adjusting the memory load to the
highest value that makes sure that only full rows are written. In this
case:
\begin{equation}
  N_\mathrm{cluster\_reads\_v1}= n + \frac{bR}{m|}d
\end{equation}
where $m|=\frac{Rb}{\sqrt[3]{n}^2}\floor*{\frac{m\sqrt[3]{n}^2}{Rb}}$

And if $m\geq\frac{Rb}{\sqrt[3]{n}}$, i.e., a full slice of blocks
fits in memory, then a second variant adjusts
the memory load to the highest value that makes sure that only full
slices are written. In this case:
\begin{equation}
  N_\mathrm{cluster\_reads\_v2}= n + \frac{bR}{m||}d
\end{equation}
where $m||=\frac{Rb}{\sqrt[3]{n}}\floor*{\frac{m\sqrt[3]{n}}{Rb}}$

@@

\todo{could we merge without de-compressing?}
